---
title: "Regression Analysis: Companion R Notebook"
subtitle: "Version 0.9"
author: "Alexis Diamond, PhD (assisted by Claude and ChatGPT)"
date: "August 26, 2026"
output: html_document
editor_options:
  chunk_output_type: console
---

This notebook accompanies the regression-analysis chapters. Run the chunks in order. All statistical and plotting code uses base R grammar; no tidyverse syntax is required.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Chapter 1: Introducing Regression as a Prediction Machine

## Conditional means and a fitted regression

The empirical conditional means show what it means to ask for the expected value of `Y` at each observed value of `X`. The fitted line is one rule for generating predictions.

Read "expected value" as *average*, not as the value we anticipate seeing. The two chunks below make that distinction concrete.

```{r ch1-conditional-means}
x <- rep(c(1, 2, 3, 4), each = 5)
y <- c(3, 4, 4, 5, 4,
       5, 5, 6, 7, 7,
       7, 8, 8, 9, 8,
       9, 10, 11, 10, 10)

dat <- data.frame(x, y)

tapply(dat$y, dat$x, mean)

fit <- lm(y ~ x, data = dat)
summary(fit)

plot(
  dat$x,
  dat$y,
  xlab = "X",
  ylab = "Y",
  main = "Observed Outcomes and a Fitted Regression"
)
abline(fit, lwd = 2)
```

## An expected value need not be observable

The word *expect* is a technical term here. An expected value is the center of a
distribution, and the center is often a value that no individual could ever have.

```{r ch1-expectation-not-observable}
# A binary outcome: 1 = employed, 0 = not employed.
employed <- c(1, 1, 1, 1, 0, 1, 1, 1, 0, 1)

mean(employed)   # 0.8

# Every person is a 0 or a 1, so no one has the value 0.8:
sort(unique(employed))

# For a binary Y the conditional mean IS the probability that Y = 1.
# A count outcome makes the same point:
visits <- c(7, 9, 10, 11, 14)
mean(visits)     # 10.2 -- not a possible number of visits
```

The conditional mean is still the right prediction in both cases. It is the best
single guess about the group, not a forecast of any one person's outcome.

## Linear regression and LOESS

Both methods are prediction machines for a conditional mean, but they impose very different amounts of structure.

```{r ch1-linear-loess}
set.seed(123)

x <- seq(0, 10, length.out = 60)
y <- 2 + 0.35 * x + 1.5 * sin(x) + rnorm(60, sd = 0.5)

dat_curve <- data.frame(x, y)

fit_linear <- lm(y ~ x, data = dat_curve)
fit_loess <- loess(y ~ x, data = dat_curve, span = 0.60)

plot(
  dat_curve$x,
  dat_curve$y,
  xlab = "X",
  ylab = "Y",
  main = "Linear Regression and LOESS"
)
abline(fit_linear, lwd = 2)

ord <- order(dat_curve$x)
lines(
  dat_curve$x[ord],
  predict(fit_loess)[ord],
  lwd = 2,
  lty = 2
)

legend(
  "topleft",
  legend = c("Linear regression", "LOESS"),
  lty = c(1, 2),
  lwd = 2,
  bty = "n"
)
```

## Different regression families for different outcomes

The following examples show continuous, binary, count, and flexible conditional-mean models.

```{r ch1-regression-families}
set.seed(42)

# Continuous outcome
x_cont <- runif(200, 0, 10)
y_cont <- 4 + 2 * x_cont + rnorm(200, sd = 3)
dat_cont <- data.frame(x = x_cont, y_cont = y_cont)
fit_ols <- lm(y_cont ~ x, data = dat_cont)

# Binary outcome
x_binary <- runif(200, 0, 10)
p_binary <- 1 / (1 + exp(-(-3 + 0.6 * x_binary)))
y_binary <- rbinom(200, size = 1, prob = p_binary)
dat_binary <- data.frame(x = x_binary, y_binary = y_binary)
fit_logit <- glm(
  y_binary ~ x,
  family = binomial(link = "logit"),
  data = dat_binary
)

# Count outcome
x_count <- runif(200, 0, 5)
lambda <- exp(0.4 + 0.25 * x_count)
count <- rpois(200, lambda = lambda)
dat_count <- data.frame(x = x_count, count = count)
fit_pois <- glm(
  count ~ x,
  family = poisson(link = "log"),
  data = dat_count
)

# Flexible conditional-mean smoother
fit_loess <- loess(y ~ x, data = dat_curve)

summary(fit_ols)
summary(fit_logit)
summary(fit_pois)
```

## Zeros in a count regression

An observed count can be zero. A fitted Poisson mean never is.

```{r ch1-count-zeros}
# Zeros are ordinary observations in count data:
sum(dat_count$count == 0)
min(dat_count$count)

# But no fitted value is ever zero, because lambda = exp(X %*% beta) > 0 always:
lambda_hat <- fitted(fit_pois)
min(lambda_hat)

# The model never predicts zero; it assigns zero a probability.
# At the smallest fitted mean, P(Y = 0) = exp(-lambda):
exp(-min(lambda_hat))
```

# Chapter 2: Technical Foundations of OLS

## Binary predictor: coefficient equals the difference in means

```{r ch2-binary-difference-means}
D <- c(rep(0, 5), rep(1, 5))
Y <- c(4, 5, 6, 7, 8,
       7, 8, 9, 10, 11)

dat_binary_ols <- data.frame(D, Y)

fit_binary <- lm(Y ~ D, data = dat_binary_ols)
summary(fit_binary)
coef(fit_binary)

mean(dat_binary_ols$Y[dat_binary_ols$D == 0])
mean(dat_binary_ols$Y[dat_binary_ols$D == 1])
mean(dat_binary_ols$Y[dat_binary_ols$D == 1]) -
  mean(dat_binary_ols$Y[dat_binary_ols$D == 0])
```

```{r ch2-binary-plot}
plot(
  jitter(dat_binary_ols$D, amount = 0.04),
  dat_binary_ols$Y,
  xaxt = "n",
  xlab = "Binary predictor D",
  ylab = "Outcome Y",
  main = "OLS with a Binary Predictor"
)
axis(1, at = c(0, 1), labels = c("D = 0", "D = 1"))

abline(fit_binary, lwd = 2)
points(
  c(0, 1),
  c(
    mean(dat_binary_ols$Y[dat_binary_ols$D == 0]),
    mean(dat_binary_ols$Y[dat_binary_ols$D == 1])
  ),
  pch = 15,
  cex = 1.5
)
```

## Regression coefficients are literal prediction instructions

```{r ch2-cars-predict}
fit_cars <- lm(dist ~ speed, data = cars)
summary(fit_cars)
coef(fit_cars)

# Predictions for training observations
train_predictions <- predict(fit_cars)
head(data.frame(
  speed = cars$speed,
  observed_dist = cars$dist,
  predicted_dist = train_predictions
))

# Predictions for new cases
new_cases <- data.frame(speed = c(10, 15, 20))
new_cases$predicted_dist <- predict(
  fit_cars,
  newdata = new_cases
)
new_cases

# Reproduce predict() literally from the coefficients
b <- coef(fit_cars)
manual_predictions <- b[1] + b[2] * new_cases$speed
cbind(new_cases, manual_predictions = manual_predictions)
```

```{r ch2-cars-prediction-plot}
plot(
  cars$speed,
  cars$dist,
  xlab = "Speed",
  ylab = "Stopping distance",
  main = "Regression as a Prediction Rule"
)
abline(fit_cars, lwd = 2)
points(
  new_cases$speed,
  new_cases$predicted_dist,
  pch = 15,
  cex = 1.4
)
```

## Coefficient uncertainty: standard errors, p-values, and confidence intervals

```{r ch2-coefficient-ci}
confint(fit_cars, level = 0.95)

coef_table <- summary(fit_cars)$coefficients
coef_table

b_speed <- coef_table["speed", "Estimate"]
se_speed <- coef_table["speed", "Std. Error"]

# Approximate 95% confidence interval
b_speed + c(-1, 1) * 2 * se_speed
```

## Mean squared error and R-squared

```{r ch2-rsquared}
fit_cars <- lm(dist ~ speed, data = cars)

summary(fit_cars)$r.squared

mse_mean_only <- mean(
  (cars$dist - mean(cars$dist))^2
)

mse_model <- mean(
  (cars$dist - predict(fit_cars))^2
)

mse_mean_only
mse_model
1 - mse_model / mse_mean_only
```

## Reducible and irreducible error

Here we know the data-generating process, so we can measure the floor that no
model can get under. Reducible error shrinks as the training sample grows;
irreducible error does not move at all.

```{r ch2-irreducible-error}
set.seed(7)

# The true DGP: Y = f(X) + epsilon, with f known to us only in a simulation.
f <- function(x) 2 + 0.8 * x
sigma <- 2

simulate_data <- function(n) {
  x <- runif(n, 0, 10)
  data.frame(x = x, y = f(x) + rnorm(n, sd = sigma))
}

# A large test set, used to measure out-of-sample prediction error.
test <- simulate_data(5000)

# The oracle: predict using the TRUE f. No model can beat this.
oracle_mse <- mean((test$y - f(test$x))^2)

# Fit on training samples of different sizes and score on the same test set.
test_mse_at <- function(n) {
  train <- simulate_data(n)
  fit <- lm(y ~ x, data = train)
  mean((test$y - predict(fit, newdata = test))^2)
}

round(
  c(
    irreducible = sigma^2,      # Var(epsilon): the floor
    oracle      = oracle_mse,   # the true f, estimated on nothing
    n_25        = test_mse_at(25),
    n_250       = test_mse_at(250),
    n_10000     = test_mse_at(10000)
  ),
  3
)
```

The fitted models approach the oracle as `n` grows -- that closing gap is the
reducible error. What none of them get below is `sigma^2`.

# Chapter 3: Prediction Uncertainty and Multiple Regression

## Three different interval targets

```{r ch3-three-intervals}
fit <- lm(dist ~ speed, data = cars)

# 1. Confidence intervals for regression coefficients
confint(fit)

# 2. Confidence interval for the conditional mean at speed = 15
predict(
  fit,
  newdata = data.frame(speed = 15),
  interval = "confidence"
)

# 3. Prediction interval for one new observation at speed = 15
predict(
  fit,
  newdata = data.frame(speed = 15),
  interval = "prediction"
)
```

## Why a coefficient's interval cannot be rescaled into a prediction's

The estimated coefficients are not independent, so uncertainty about a fitted
value depends on the whole variance-covariance matrix, not on one standard
error.

```{r ch3-vcov-naive-first-difference}
fit <- lm(dist ~ speed, data = cars)

# The variance-covariance matrix: variances on the diagonal,
# covariances off it.
vcov(fit)

# As a correlation, the two estimates are nearly redundant:
cov2cor(vcov(fit))

# The off-diagonal entry is exactly -xbar * Var(beta1hat):
c(
  off_diagonal = vcov(fit)["(Intercept)", "speed"],
  formula      = -mean(cars$speed) * vcov(fit)["speed", "speed"]
)

# CORRECT interval for the conditional mean at speed = 15:
predict(fit, newdata = data.frame(speed = 15), interval = "confidence")

# NAIVE first difference: hold the intercept fixed and swing the slope
# across the ends of its own confidence interval.
coef(fit)[1] + confint(fit)["speed", ] * 15
```

The naive interval is about three times too wide. The decomposition shows why:
the covariance term is negative and larger in magnitude than either variance,
so most of the coefficient uncertainty cancels when the two are combined.

```{r ch3-variance-decomposition}
V <- vcov(fit)
x <- 15

terms <- c(
  var_b0     = V[1, 1],
  x2_var_b1  = x^2 * V[2, 2],
  cross_term = 2 * x * V[1, 2]
)
terms
sum(terms)

# Standard errors: correct versus naive.
c(
  correct = sqrt(sum(terms)),
  naive   = x * sqrt(V[2, 2])
)

# The general formula, c' V c, reproduces what predict() reports:
cvec <- c(1, x)
sqrt(t(cvec) %*% V %*% cvec)
```

With no intercept there is only one coefficient, so there is no covariance to
ignore and the shortcut becomes exact.

```{r ch3-no-intercept-exact}
fit0 <- lm(dist ~ speed - 1, data = cars)

vcov(fit0)   # a 1x1 matrix

predict(fit0, newdata = data.frame(speed = 15), interval = "confidence")

confint(fit0)["speed", ] * 15   # identical, to every printed digit
```

The naive width is a straight line in `x`; the correct width is a hyperbola
with its minimum at `xbar`. Different shapes, so they can cross but cannot
coincide. Note the `x = 0` row: the shortcut claims perfect certainty about
`E[Y | X = 0]`, which is just the intercept, whose standard error is printed in
the regression output.

```{r ch3-width-comparison}
tq <- qt(0.975, df.residual(fit))
s  <- summary(fit)$sigma

naive_width <- function(x) 2 * tq * x * sqrt(V[2, 2])
ci_width    <- function(x) 2 * tq * sqrt(V[1, 1] + x^2 * V[2, 2] + 2 * x * V[1, 2])
pi_width    <- function(x) 2 * tq * sqrt(s^2 + V[1, 1] + x^2 * V[2, 2] + 2 * x * V[1, 2])

# The one x where the shortcut happens to be right:
# solve Var(b0) + 2x Cov = 0  =>  x* = mean(x^2) / (2 * mean(x))
xstar <- mean(cars$speed^2) / (2 * mean(cars$speed))

xs <- c(0, 5, xstar, mean(cars$speed), 20, 25, 100)
round(
  data.frame(
    x             = xs,
    naive         = naive_width(xs),
    correct_ci    = ci_width(xs),
    prediction    = pi_width(xs),
    naive_over_ci = naive_width(xs) / ci_width(xs)
  ),
  3
)

range(cars$speed)   # x* falls INSIDE the observed data
```

The shortcut is too narrow for small `x`, exactly right at one point, and
almost three times too wide near the middle of the data. It is not reliably
conservative, and it never comes close to the prediction interval -- it
contains no information about the spread of `Y` around the line at all.

A first difference across two values of `x` is a different question. In the
simple model the intercept cancels, so rescaling the slope's interval is
legitimate -- but that survives only because the difference loads on a single
coefficient. Add a quadratic term and it fails again.

```{r ch3-first-difference-boundary}
# Simple model: the difference depends on beta1 alone, so both agree.
c(
  full  = sqrt(t(c(0, 5)) %*% vcov(fit) %*% c(0, 5)),
  naive = 5 * sqrt(vcov(fit)["speed", "speed"])
)

# Quadratic model: the same difference now loads on TWO coefficients.
fitq <- lm(dist ~ speed + I(speed^2), data = cars)
cq <- c(0, 20 - 15, 20^2 - 15^2)

c(
  full  = sqrt(t(cq) %*% vcov(fitq) %*% cq),
  naive = 5 * sqrt(vcov(fitq)["speed", "speed"]),
  corr  = cov2cor(vcov(fitq))["speed", "I(speed^2)"]
)
```

## Visualizing a coefficient confidence interval

```{r ch3-coefficient-ci-plot}
fit <- lm(dist ~ speed, data = cars)

slope <- coef(fit)["speed"]
slope_ci <- confint(fit)["speed", ]

plot(
  1, slope,
  xlim = c(0.5, 1.5),
  ylim = range(c(0, slope_ci)),
  xaxt = "n",
  xlab = "",
  ylab = "Slope coefficient for speed",
  pch = 19
)

segments(
  x0 = 1,
  y0 = slope_ci[1],
  x1 = 1,
  y1 = slope_ci[2],
  lwd = 3
)

points(1, slope, pch = 19)
axis(1, at = 1, labels = "speed")
abline(h = 0, lty = 3)
```

## Plot confidence and prediction bands

```{r ch3-interval-bands}
fit <- lm(dist ~ speed, data = cars)

speed_grid <- data.frame(
  speed = seq(
    from = min(cars$speed),
    to = max(cars$speed),
    length.out = 200
  )
)

conf_band <- predict(
  fit,
  newdata = speed_grid,
  interval = "confidence",
  level = 0.95
)

pred_band <- predict(
  fit,
  newdata = speed_grid,
  interval = "prediction",
  level = 0.95
)

plot(
  cars$speed,
  cars$dist,
  xlab = "Speed",
  ylab = "Stopping distance",
  main = "Confidence and Prediction Intervals"
)

lines(speed_grid$speed, conf_band[, "fit"], lwd = 3)
lines(speed_grid$speed, conf_band[, "lwr"], lty = 2, lwd = 2)
lines(speed_grid$speed, conf_band[, "upr"], lty = 2, lwd = 2)
lines(speed_grid$speed, pred_band[, "lwr"], lty = 3, lwd = 2)
lines(speed_grid$speed, pred_band[, "upr"], lty = 3, lwd = 2)

legend(
  "topleft",
  legend = c(
    "Regression line",
    "Confidence interval for mean",
    "Prediction interval for new observation"
  ),
  lty = c(1, 2, 3),
  lwd = c(3, 2, 2),
  bty = "n"
)
```

## The same point prediction sits at the center of both intervals

```{r ch3-same-center}
fit <- lm(dist ~ speed, data = cars)

b <- coef(fit)

# Calculate the prediction literally from the coefficients
manual_prediction <- b[1] + b[2] * 15
manual_prediction

# Ask predict() for the point prediction
predict(
  fit,
  newdata = data.frame(speed = 15)
)
```

## Multiple regression and holding predictors fixed

```{r ch3-mtcars-multiple}
fit_simple <- lm(mpg ~ am, data = mtcars)
fit_multiple <- lm(mpg ~ wt + am, data = mtcars)

coef(fit_simple)
coef(fit_multiple)

# Two cars with exactly the same weight
same_weight <- data.frame(
  wt = c(3, 3),
  am = c(0, 1)
)

predicted_mpg <- predict(
  fit_multiple,
  newdata = same_weight
)

predicted_mpg
predicted_mpg[2] - predicted_mpg[1]
coef(fit_multiple)["am"]

# Two cars with the same transmission, different weight
same_transmission <- data.frame(
  wt = c(3, 4),
  am = c(0, 0)
)

predicted_mpg <- predict(
  fit_multiple,
  newdata = same_transmission
)

predicted_mpg[2] - predicted_mpg[1]
coef(fit_multiple)["wt"]
```

```{r ch3-multiple-plot}
point_type <- ifelse(mtcars$am == 1, 19, 1)

plot(
  mtcars$wt,
  mtcars$mpg,
  pch = point_type,
  xlab = "Weight (1,000 pounds)",
  ylab = "Miles per gallon",
  main = "Multiple Regression: mpg ~ wt + am"
)

b <- coef(fit_multiple)
abline(a = b["(Intercept)"], b = b["wt"], lwd = 3)
abline(
  a = b["(Intercept)"] + b["am"],
  b = b["wt"],
  lwd = 3,
  lty = 2
)

legend(
  "topright",
  legend = c("Automatic: am = 0", "Manual: am = 1"),
  pch = c(1, 19),
  lty = c(1, 2),
  lwd = c(3, 3),
  bty = "n"
)
```

## Interactions and centering

```{r ch3-interaction-centering}
fit_interaction <- lm(mpg ~ wt * am, data = mtcars)
summary(fit_interaction)

mtcars_for_example <- mtcars
mtcars_for_example$wt_centered <-
  mtcars_for_example$wt - mean(mtcars_for_example$wt)

fit_interaction_centered <- lm(
  mpg ~ wt_centered * am,
  data = mtcars_for_example
)
summary(fit_interaction_centered)
```

## A confidence interval when a confounder is omitted

The short regression omits the confounder `z`. Its confidence interval is still
a valid 95% interval -- for the projection coefficient `beta1`, not for the
causal effect `alpha1`. Watch what happens to each coverage rate as `n` grows.

```{r ch3-omitted-confounder-coverage}
set.seed(11)

alpha1 <- 2.00   # the causal effect of x on y
alpha2 <- 0.15   # the effect of the omitted confounder z on y

one_study <- function(n) {
  z <- rnorm(n)
  x <- 0.7 * z + rnorm(n)                      # x and z are related
  y <- 1 + alpha1 * x + alpha2 * z + rnorm(n)
  fit <- lm(y ~ x)                             # z omitted
  ci <- confint(fit)["x", ]
  c(est = unname(coef(fit)["x"]),
    lower = unname(ci[1]),
    upper = unname(ci[2]))
}

# What the short regression actually targets: beta1 = alpha1 + alpha2 * delta1,
# where delta1 is the population slope of z on x.
delta1 <- 0.7 / (0.7^2 + 1)
beta1 <- alpha1 + alpha2 * delta1
beta1

coverage <- function(n, reps = 2000) {
  out <- replicate(reps, one_study(n))
  c(n                 = n,
    mean_estimate     = mean(out["est", ]),
    covers_causal     = mean(out["lower", ] <= alpha1 & out["upper", ] >= alpha1),
    covers_projection = mean(out["lower", ] <= beta1  & out["upper", ] >= beta1))
}

round(rbind(coverage(100), coverage(1000), coverage(10000)), 4)
```

Coverage of `beta1` sits at the advertised 95% at every sample size. Coverage of
`alpha1` falls toward zero, because the interval narrows while the bias does
not. The estimates are not getting worse -- they are converging, precisely, on
the wrong quantity.

# Chapter 4: Functional Form, Interactions, Interpolation, and Extrapolation

## Functional form

```{r ch4-functional-form}
set.seed(9)

x <- seq(1, 10, length.out = 100)
y <- 5 + 2 * x - 0.25 * x^2 + rnorm(100, sd = 1.5)
dat_functional <- data.frame(x, y)

fit_linear <- lm(y ~ x, data = dat_functional)
fit_quadratic <- lm(y ~ x + I(x^2), data = dat_functional)
fit_cubic <- lm(y ~ x + I(x^2) + I(x^3), data = dat_functional)
fit_logx <- lm(y ~ log(x), data = dat_functional)

summary(fit_linear)$r.squared
summary(fit_quadratic)$r.squared
summary(fit_cubic)$r.squared
summary(fit_logx)$r.squared
```

## Additive model versus interaction model

```{r ch4-mtcars-interaction}
fit_additive <- lm(mpg ~ wt + am, data = mtcars)
fit_interaction <- lm(mpg ~ wt * am, data = mtcars)

summary(fit_additive)
summary(fit_interaction)
```

```{r ch4-interaction-plot}
plot(
  mtcars$wt,
  mtcars$mpg,
  pch = ifelse(mtcars$am == 0, 1, 19),
  xlab = "Weight (1,000 pounds)",
  ylab = "Miles per gallon",
  main = "Interaction: mpg ~ wt * am"
)

auto_wt <- seq(
  min(mtcars$wt[mtcars$am == 0]),
  max(mtcars$wt[mtcars$am == 0]),
  length.out = 100
)

manual_wt <- seq(
  min(mtcars$wt[mtcars$am == 1]),
  max(mtcars$wt[mtcars$am == 1]),
  length.out = 100
)

auto_pred <- predict(
  fit_interaction,
  newdata = data.frame(wt = auto_wt, am = 0)
)

manual_pred <- predict(
  fit_interaction,
  newdata = data.frame(wt = manual_wt, am = 1)
)

lines(auto_wt, auto_pred, lwd = 3)
lines(manual_wt, manual_pred, lwd = 3, lty = 2)

legend(
  "topright",
  legend = c("Automatic", "Manual"),
  pch = c(1, 19),
  lty = c(1, 2),
  lwd = c(3, 3),
  bty = "n"
)
```

## Interpolation across an unsupported interior gap

```{r ch4-gap-data}
set.seed(26)

x_left <- runif(35, min = 1, max = 4)
x_right <- runif(35, min = 7, max = 10)
x <- c(x_left, x_right)

y <- c(
  2 + 3 * x_left + rnorm(35, mean = 0, sd = 0.5),
  28 - x_right + rnorm(35, mean = 0, sd = 0.5)
)

gap_data <- data.frame(x = x, y = y)

fit_linear <- lm(y ~ x, data = gap_data)
fit_quadratic <- lm(y ~ x + I(x^2), data = gap_data)

new_point <- data.frame(x = 5.5)
predict(fit_linear, newdata = new_point)
predict(fit_quadratic, newdata = new_point)
```

```{r ch4-gap-plot}
x_grid <- seq(1, 10, length.out = 300)
new_grid <- data.frame(x = x_grid)

linear_hat <- predict(fit_linear, newdata = new_grid)
quadratic_hat <- predict(fit_quadratic, newdata = new_grid)

plot(
  gap_data$x,
  gap_data$y,
  xlab = "X",
  ylab = "Y",
  main = "Interpolation Across a Gap in the Training Data",
  pch = 19
)

rect(
  xleft = 4,
  ybottom = par("usr")[3],
  xright = 7,
  ytop = par("usr")[4],
  density = 12,
  angle = 45,
  border = NA
)

points(gap_data$x, gap_data$y, pch = 19)
lines(x_grid, linear_hat, lwd = 3)
lines(x_grid, quadratic_hat, lwd = 3, lty = 2)
abline(v = c(4, 7), lty = 3)

points(
  5.5,
  predict(fit_linear, newdata = data.frame(x = 5.5)),
  pch = 1,
  cex = 1.5,
  lwd = 2
)

points(
  5.5,
  predict(fit_quadratic, newdata = data.frame(x = 5.5)),
  pch = 2,
  cex = 1.5,
  lwd = 2
)

legend(
  "topleft",
  legend = c(
    "Linear fit",
    "Quadratic fit",
    "Gap with no training observations"
  ),
  lty = c(1, 2, NA),
  lwd = c(3, 3, NA),
  fill = c(NA, NA, "black"),
  density = c(0, 0, 35),
  angle = 45,
  border = c(NA, NA, "black"),
  bty = "n"
)
```

## Extrapolation and functional-form dependence with `faithful`

```{r ch4-faithful-models}
fit_linear <- lm(waiting ~ eruptions, data = faithful)
fit_quadratic <- lm(
  waiting ~ eruptions + I(eruptions^2),
  data = faithful
)
fit_cubic <- lm(
  waiting ~ eruptions + I(eruptions^2) + I(eruptions^3),
  data = faithful
)

summary(fit_linear)$r.squared
summary(fit_quadratic)$r.squared
summary(fit_cubic)$r.squared

new_case <- data.frame(eruptions = 7)
predict(fit_linear, newdata = new_case)
predict(fit_quadratic, newdata = new_case)
predict(fit_cubic, newdata = new_case)
```

```{r ch4-faithful-plot}
x_grid <- seq(1.4, 7, length.out = 300)
new_grid <- data.frame(eruptions = x_grid)

plot(
  faithful$eruptions,
  faithful$waiting,
  xlim = c(1.4, 7),
  ylim = c(15, 115),
  xlab = "Eruption duration (minutes)",
  ylab = "Waiting time (minutes)",
  main = "Functional Form and Extrapolation"
)

lines(x_grid, predict(fit_linear, newdata = new_grid), lwd = 3)
lines(
  x_grid,
  predict(fit_quadratic, newdata = new_grid),
  lwd = 3,
  lty = 2
)
lines(
  x_grid,
  predict(fit_cubic, newdata = new_grid),
  lwd = 3,
  lty = 3
)

abline(v = max(faithful$eruptions), lty = 2)

legend(
  "topleft",
  legend = c("Linear", "Quadratic", "Cubic"),
  lty = c(1, 2, 3),
  lwd = c(3, 3, 3),
  bty = "n"
)
```

## LOESS and extrapolation

```{r ch4-loess-extrapolation}
fit_loess <- loess(waiting ~ eruptions, data = faithful)

predict(
  fit_loess,
  newdata = data.frame(eruptions = c(2, 4, 7))
)
```

## Support in multiple regression: an unequally supported comparison

Automatic cars in `mtcars` weigh roughly 2.47 to 5.42 thousand pounds; manual
cars roughly 1.51 to 3.57. At a weight of 5, the automatic prediction is
supported by nearby observations, but the manual prediction is an extrapolation.

```{r ch4-support-comparison}
fit_interaction <- lm(mpg ~ wt * am, data = mtcars)

predict(
  fit_interaction,
  newdata = data.frame(
    wt = c(5, 5),
    am = c(0, 1)
  )
)

range(mtcars$wt[mtcars$am == 0])
range(mtcars$wt[mtcars$am == 1])
```

# Chapter 5: Regression, Overlap, Model Dependence, and Causal Inference

## A constructed observational study with limited overlap

The treatment effect in this constructed dataset is exactly 6. Treatment status is strongly related to `x`, and the untreated outcome function is nonlinear.

```{r ch5-observational-data}
x_control <- rep(0:8, each = 3)
x_treated <- rep(5:16, each = 3)

noise_control <- rep(c(-1, 0, 1), times = 9)
noise_treated <- rep(c(-1, 0, 1), times = 12)

Y0_control <- 5 + x_control + 0.4 * x_control^2 + noise_control
Y0_treated <- 5 + x_treated + 0.4 * x_treated^2 + noise_treated

D <- c(
  rep(0, length(x_control)),
  rep(1, length(x_treated))
)

x <- c(x_control, x_treated)
Y <- c(Y0_control, Y0_treated + 6)

causal_dat <- data.frame(Y, D, x)
```

## Model dependence under poor overlap

```{r ch5-model-dependence}
fit_linear_adjustment <- lm(
  Y ~ D + x,
  data = causal_dat
)

fit_quadratic_adjustment <- lm(
  Y ~ D + x + I(x^2),
  data = causal_dat
)

coef(fit_linear_adjustment)["D"]
coef(fit_quadratic_adjustment)["D"]
```

The two specifications use the same observations but impose different functional forms. The correctly specified quadratic model recovers 6, while the linear adjustment gives a very different answer because the groups have limited overlap.

## Restrict to the region of common support

```{r ch5-overlap-restriction}
overlap_dat <- causal_dat[
  causal_dat$x >= 5 & causal_dat$x <= 8,
]

fit_linear_overlap <- lm(
  Y ~ D + x,
  data = overlap_dat
)

fit_quadratic_overlap <- lm(
  Y ~ D + x + I(x^2),
  data = overlap_dat
)

coef(fit_linear_overlap)["D"]
coef(fit_quadratic_overlap)["D"]
```

In this deliberately clean example, both specifications now return 6 because treated and control outcomes are directly compared at the same `x` values.

```{r ch5-overlap-plot}
plot(
  causal_dat$x,
  causal_dat$Y,
  pch = ifelse(causal_dat$D == 1, 19, 1),
  xlab = "Confounder X",
  ylab = "Observed outcome Y",
  main = "Limited Overlap in an Observational Study"
)

abline(v = 5, lty = 3)
abline(v = 8, lty = 3)

legend(
  "topleft",
  legend = c("Control", "Treated", "Overlap: X = 5 to 8"),
  pch = c(1, 19, NA),
  lty = c(NA, NA, 3),
  bty = "n"
)
```

## A perfectly balanced randomized-design illustration

This final example represents a perfectly balanced assignment within each value of `x`: three of six units receive treatment and three receive control. The true treatment effect remains 6. Because treatment and control have the same covariate support and distribution here, the treatment-only regression and two differently specified adjusted regressions all recover the same treatment contrast in this constructed dataset.

```{r ch5-balanced-rct}
x <- rep(0:16, each = 6)
D <- rep(c(0, 0, 0, 1, 1, 1), times = 17)
noise <- rep(c(-1, 0, 1, -1, 0, 1), times = 17)

Y0 <- 5 + x + 0.4 * x^2 + noise
Y <- Y0 + 6 * D

rct_dat <- data.frame(Y, D, x)

# Difference in means = treatment-only OLS coefficient
mean(rct_dat$Y[rct_dat$D == 1]) -
  mean(rct_dat$Y[rct_dat$D == 0])

fit_rct_simple <- lm(Y ~ D, data = rct_dat)
fit_rct_linear <- lm(Y ~ D + x, data = rct_dat)
fit_rct_quadratic <- lm(Y ~ D + x + I(x^2), data = rct_dat)

coef(fit_rct_simple)["D"]
coef(fit_rct_linear)["D"]
coef(fit_rct_quadratic)["D"]
```

# Suggested experiments for students

These are useful modifications to try after working through the notebook:

1. Change the values passed to `predict()` in the `cars` examples and watch how point predictions change.
2. Compare confidence and prediction interval widths near the middle and edges of the observed `speed` range.
3. Add an irrelevant predictor to an OLS model and verify that training-set R-squared cannot fall.
4. In the interpolation-gap example, move the gap or make it wider and compare the linear and quadratic predictions.
5. In the causal-overlap example, gradually shrink the overlap region and watch the treatment coefficient become more sensitive to functional form.
6. In the `mtcars` interaction model, predict manual and automatic mileage at the same weights and inspect where those comparisons are supported by actual data.
